---
title: "Replicating Research with LLMs"
author: "Evan Gorstein"
date: "2025-10-30"
categories: [llms, replication]
draft: true
---


```{r}
library(tidyverse)
library(haven)
```

```{r}
fulldata <- read_dta("data/maindata.dta")
```

## Basic demographics
```{r}
demo <- fulldata %>%
  rename(
    id = R0000100, hid = R0000149, MONTH79 = R0000300, YR79 = R0000500,
    sample = R0173600, RACE79 = R0214700, SEX79 = R0214800,	MONTH81 = R0410100,
    YR81 = R0410300, SEX82 = R0810200
  ) %>%
  mutate(
    year = if_else(YR81 > 0, YR81, YR79),
    month = if_else(MONTH81 > 0, MONTH81, MONTH79)
    ) %>%
  mutate(
    race = factor(RACE79, levels = c(3, 1, 2), labels = c("white", "hisp", "black")),
    sex = factor(case_when(
      SEX82==2 ~ "female",
      SEX82 < 0 & SEX79==2 ~ "female",
      .default = "male"
    ), levels = c("male", "female"))
    ) %>%
  mutate(
    sample = case_when(
      sample >= 1 & sample <= 8 ~ "cross-sec",
      sample >= 9 & sample <= 14 ~ "supplemental",
      sample >= 15 & sample <= 20 ~ "military"
    )
    )
# check that all sample types are covered
any(is.na(demo$sample))
```

```{r}
demo <- demo %>%
  select(id, hid, year, month, sex, race, sample) %>% 
  arrange(id)
```


## Highest grade completed
```{r}
# --- Stata: rename R*... to hgc*... ---
hgc <- fulldata %>%
  rename(
    id = R0000100,
    hgc79 = R0216701, hgc80 = R0406401, hgc81 = R0618901, hgc82 = R0898201,
    hgc83 = R1145001, hgc84 = R1520201, hgc85 = R1890901, hgc86 = R2258001,
    hgc87 = R2445401, hgc88 = R2871101, hgc89 = R3074801, hgc90 = R3401501,
    hgc91 = R3656901, hgc92 = R4007401, hgc93 = R4418501, hgc94 = R5103900,
    hgc96 = R5166901, hgc98 = R6479600, hgc100 = R7007300, hgc102 = R7704400,
    hgc104 = R8496800
  )

# --- Stata: keep hgc* id; ---
hgc <- hgc %>%
  select(id, starts_with("hgc"))
```


```{r}
hgc <- hgc %>%
  pivot_longer(
    cols = starts_with("hgc"),
    names_to = "year",
    values_to = "hgc",
    names_prefix = "hgc", # Removes "hgc" from "hgc79" to get "79"
    names_transform = list(year = as.integer) # Converts "79" to 79
  ) %>%
  mutate(
    # gen below6=0; replace below6=1 if hgc<6 & hgc>=0;
    # We use if_else(condition, true, false, missing)
    # Stata's `gen below6=0` sets NA rows to 0, so we use `missing = 0`.
    below6 = if_else(hgc < 6 & hgc >= 0, 1, 0, 0),
    
    # replace hgc=6 if hgc<6 & hgc>=0;
    # replace hgc=6 if hgc>90;
    # case_when() handles this cleanly.
    hgc = case_when(
      hgc < 6 & hgc >= 0 ~ 6,
      hgc > 90 ~ 6,
      .default = hgc # Keeps all other values (including NA) as they are
    )
  ) %>%
  pivot_wider(
    id_cols = id,
    names_from = year,
    values_from = c(hgc, below6),
    # This creates "hgc79", "below679", etc., to match Stata's output
    names_glue = "{.value}{year}" 
  )

hgc <- hgc %>%
  mutate(
    across(
      starts_with("hgc"),
      ~ if_else(.x < 0 , NA_real_, .x)
    )
  )
```

```{r}
 final <- demo %>%
  left_join(
    hgc, by = "id"
  ) %>%
  arrange(id)
```


## Employment Status, Class of Workers, Hours worked

Employment status and class of worker: 
```{r}
employ <- fulldata %>%
  rename(
    # General ID
    id = R0000100,

    # 1979
    cls79 = R0046800, # For those who work outside the house, do you work at a private company (1), for the gov (2), self-employed (3), without pay (4)
    esr79 = R0214900, # Do you work (1), with job, not at work (2) are you unemployed (3), are you keeping house (4), are you a student (5), are you unable to work (6), Other (7), in military (8)
    curr1job79 = R0094700, # For those who recorded at least one employer, is job 1 the one that is current?
    curr2job79 = R0094800, # For those who recorded at least two employers, is job 2 the one that is current?
    curr3job79 = R0094900, # Etc.
    curr4job79 = R0095000,
    curr5job79 = R0095100,

    # 1980
    cls80 = R0263500,
    esr80 = R0406300,
    curr1job80 = R0337700,
    curr2job80 = R0349200,
    curr3job80 = R0360700,
    curr4job80 = R0372200,
    curr5job80 = R0383700,

    # 1981
    cls81 = R0446600,
    esr81 = R0618800,
    curr1job81 = R0545500,
    curr2job81 = R0558600,
    curr3job81 = R0571700,
    curr4job81 = R0584800,
    curr5job81 = R0597900,

    # 1982
    cls82 = R0702300,
    esr82 = R0898500,
    curr1job82 = R0840100,
    curr2job82 = R0853200,
    curr3job82 = R0866300,
    curr4job82 = R0879400,
    curr5job82 = R0892500,

    # 1983
    cls83 = R0945400,
    esr83 = R1144700,
    curr1job83 = R1087200,
    curr2job83 = R1100400,
    curr3job83 = R1113600,
    curr4job83 = R1126800,
    curr5job83 = R1140000,

    # 1984
    cls84 = R1255800,
    esr84 = R1519900,
    curr1job84 = R1462900,
    curr2job84 = R1476000,
    curr3job84 = R1489100,
    curr4job84 = R1502200,
    curr5job84 = R1515300,

    # 1985
    cls85 = R1650600,
    esr85 = R1890600,
    curr1job85 = R1809700,
    curr2job85 = R1822400,
    curr3job85 = R1835100,
    curr4job85 = R1847800,
    curr5job85 = R1860500,

    # 1986
    cls86 = R1923200,
    esr86 = R2257700,
    curr1job86 = R2171400,
    curr2job86 = R2185000,
    curr3job86 = R2198600,
    curr4job86 = R2212200,
    curr5job86 = R2225800,

    # 1987
    cls87 = R2318000,
    esr87 = R2445100,
    curr1job87 = R2376200,
    curr2job87 = R2387500,
    curr3job87 = R2398800,
    curr4job87 = R2410100,
    curr5job87 = R2421400,

    # 1988
    cls88 = R2525800,
    esr88 = R2870600,
    curr1job88 = R2770800,
    curr2job88 = R2783700,
    curr3job88 = R2796600,
    curr4job88 = R2809500,
    curr5job88 = R2822400,

    # 1989
    cls89 = R2924800,
    esr89 = R3074300,
    curr1job89 = R3012600,
    curr2job89 = R3025700,
    curr3job89 = R3038800,
    curr4job89 = R3051900,
    curr5job89 = R3065000,

    # 1990
    cls90 = R3127500,
    esr90 = R3401000,
    curr1job90 = R3340000,
    curr2job90 = R3354000,
    curr3job90 = R3368000,
    curr4job90 = R3382000,
    curr5job90 = R3396000,

    # 1991
    cls91 = R3523200,
    esr91 = R3656400,
    curr1job91 = R3604300,
    curr2job91 = R3616400,
    curr3job91 = R3628500,
    curr4job91 = R3640600,
    curr5job91 = R3652700,

    # 1992
    cls92 = R3728200,
    esr92 = R4006900,
    curr1job92 = R3954500,
    curr2job92 = R3966700,
    curr3job92 = R3978900,
    curr4job92 = R3991100,
    curr5job92 = R4003300,

    # 1993
    cls93 = R4182300,
    esr93 = R4418000,

    # 1994
    clsa94 = R4586700,
    clsb94 = R4586800,
    clsc94 = R4587900,
    esr94 = R5081000,

    # 1996
    clsa96 = R5268800,
    clsb96 = R5268900,
    clsc96 = R5270400,
    esr96 = R5166300,

    # 1998
    clsa98 = R5925000,
    clsb98 = R5925800,
    clsc98 = R5936600,
    esr98 = R6478900,

    # 2000
    cls100 = R6592300,
    esr100 = R6555000,

    # 2002
    cls102 = R7210100,
    esr102 = R7181300,

    # 2004
    cls104 = R7898500,
    esr104 = R7866700
  )
```

Bring in hours worked:
```{r}
employ <- employ %>%
  rename(
    hrsweekj179 = R0070500,
    hrsweekj279 = R0070600,
    hrsweekj379 = R0070700,
    hrsweekj479 = R0070800,
    hrsweekj579 = R0070900,
    hrsweek80 = R0264300,
    hrsweek81 = R0446900,
    hrsweek82 = R0702600,
    hrsweek83 = R0945700,
    hrsweek84 = R1256100,
    hrsweek85 = R1650900,
    hrsweek86 = R1923500,
    hrsweek87 = R2318300,
    hrsweek88 = R2526100,
    hrsweek89 = R2925100,
    hrsweek90 = R3127900,
    hrsweek91 = R3523600,
    hrsweek92 = R3728600,
    hrsweek93 = R4182600,
    hrsweek94 = R4540300,
    hrsweek96 = R5236400,
    hrsweek98 = R5838200,
    hrsweek100 = R6578200,
    hrsweek102 = R7192900,
    hrsweek104 = R7879600
  )

```

esr for years after 98 is just whether the worker is currently holding job 1 or not (not created by NLSY for these years)

```{r}
employ <- employ %>%
  mutate(
    esr100 = if_else(esr100 == 0, 3, esr100),
    esr102 = if_else(esr102 == 0, 3, esr102),
    esr104 = if_else(esr104 == 0, 3, esr104)
  )
```


```{r}
employ <- employ %>%
  select(!starts_with("R"))


employ <- employ %>%
  pivot_longer(
    cols = !id,
    names_to = c(".value", "year"),
    names_pattern = "(.*?)(100|102|104|\\d{2})$"
  )

# coding change between 1994 and 1998
employ <- employ %>%
  mutate(
    cls = case_when(
    year < 94 | year > 98 ~ cls,
    clsa > -3 ~ clsa,
    clsb > -3 ~ clsb,
    clsc > - 3 ~ clsc
      )
  ) %>%
  mutate(
    cls = case_when(
      year < 94 | year > 98 ~ cls,
      cls==2 | cls==3 ~ 1,
      cls==1 ~ 2,
      cls==4 ~ 3,
      cls==5 ~ 4,
      .default = cls
    )
  )
```

Now we combine `cls` and `esr` to create a variable `code`:

-9 = coding error but working 
0 = interviewed, non-milt, not working, 1 = working for pay, 2 = non-interview,
3 = military & no civilian job, 4= self-employed, 5 - working w/o pay

```{r}
employ <- employ %>%
  mutate(
    code = case_when(
      cls == 4 ~ 5, # working without pay
      cls == 3 ~ 4, # self-employed
      cls == 1 | cls == 2 ~ 1, # working for pay 
      esr == 8 ~ 3, # military
      (cls>-4 & cls<0) & (esr==1 | esr==2) ~ -9, # cls is missing, but is working according to esr
      (cls == -4) | ((cls>-4 & cls<0) & (esr>2 & esr!=8)) ~ 0, # neither working nor in the military
      esr == -5 ~ 2, # not interviewed
      .default = -99
    )
  )
```

Determine CPS job. For years after 1992, it's job 1. For years before 1993, it's based on currxjob variables

```{r}
employ <- employ %>%
  mutate(
    CPSJOB = case_when(
      year > 92 ~ 1, 
      curr1job == 1 ~ 1,
      curr2job == 1 ~ 2,
      curr3job == 1 ~ 3,
      curr4job == 1 ~ 4,
      curr5job == 1 ~ 5,
      .default = 0 
    )
  ) %>%
  select(!starts_with("curr"))
```

Determine how many hours per week are worked for 1979

```{r}
employ <- employ %>%
  mutate(
    hrsweek = case_when(
      year != 79 ~ hrsweek,
      CPSJOB == 5 ~ hrsweekj5,
      CPSJOB == 4 ~ hrsweekj4,
      CPSJOB == 3 ~ hrsweekj3,
      CPSJOB == 2 ~ hrsweekj2,
      CPSJOB == 1 ~ hrsweekj1,
      .default = hrsweek
    )
  )
```


Reshape back to wide and merge with fulldata
```{r}
employ <- employ %>%
  select(id, code, cls, esr, year, hrsweek) %>%
  pivot_wider(
    names_from = year,
    values_from = -c(id, year)
  ) 


final <- final %>%
  left_join(employ, by="id") %>%
  arrange(id)
```

## More stuff

Start with hourly rate of pay
```{r}
# more_stuff <- fulldata %>%
#   rename(
#     id = R0000100, 
#     hrp79 = R0047010, 
#     hrp80 = R0263710, 
#     hrp81 = ,
#     hrp82 = ,
#     hrp83 = ,
#     hrp84 = ,
#     hrp85 = ,
#     hrp86 = ,
#     hrp87 = ,
#     hrp88 = 
#   )
  
  
```










